---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r readdata, cache=TRUE}
  #setwd("E:/Development/R/repData/RepData_PeerAssessment1")
  if (!file.exists("activity"))
  {
    dir.create("activity")
    if (file.exists("activity.zip"))
    {
      message("Unzipping dataset")
      unzip("activity.zip", exdir= "./activity")
    }
  }
  stepData <- read.csv("activity/activity.csv")
  stepData$date <- as.Date(stepData$date)
  
  names(stepData) <- c("Steps","Date","Interval")
```

## What is mean total number of steps taken per day?
```{r aggregatestepsperday, echo =FALSE}
  
stepsByDay <- aggregate(x = stepData$Steps , by = list(stepData$Date), FUN = sum ,na.rm=TRUE)
names(stepsByDay) <- c("Date","Steps")
hist(stepsByDay$Steps, xlab="number of steps", main="Total number of steps taken each day",col="gray")

```

* The mean total number of steps taken per day is `r round(mean(stepsByDay$Steps),digits=4)` steps.  
* The median total number of steps taken per day is `r median(stepsByDay$Steps)` steps.

###Daily Mean

```{r dailymean, echo =TRUE}
#split data by date
dailySteps <- split(stepData$Steps,stepData$Date)
dailyMean <- sapply(dailySteps,FUN=mean)
# Convert to DF for formatting and printing
meanDF <- data.frame(Date= names(dailyMean), mean = dailyMean,row.names = NULL)
names(meanDF) <- c("Date","Mean")
print(meanDF, digits=4)
```

###Daily Median
```{r dailymedian, echo =TRUE}
dailyMeadian <- sapply(dailySteps,FUN=median)
medianDF <- data.frame(Date= names(dailyMeadian), median = dailyMeadian,row.names = NULL)
names(medianDF) <- c("Date","Median")
print(medianDF, digits=4)

```


## What is the average daily activity pattern?

```{r averageactivitypatter}

intervalsplit <- split(stepData$Steps,stepData$Interval)
meanStepsInterval <- sapply(intervalsplit, mean,na.rm=TRUE)
plot(names(meanStepsInterval), meanStepsInterval , type = "l", 
main = "Mean steps per interval for all days",  xlab = "Interval", ylab = "Mean # of steps "  )
#Find the Maximum Value and interval
maxIntervalValue <- max(meanStepsInterval, na.rm = TRUE)
maxInterval <- names(meanStepsInterval)[which.max(meanStepsInterval)]

```
The 5-minute interval (on average) occurs at `r maxInterval ` and contains  `r round(maxIntervalValue,digits=4)` steps


## Imputing missing values

The number of missing values is `r sum(is.na(stepData$Steps))`

Missing values will be replaced by the mean value for that 5-minute interval

```{r imputmissingvalues }
stepsByDayMean <- aggregate(Steps ~ Interval, FUN=mean,data=stepData)
stepsMerged <- merge (x=stepData,y=stepsByDayMean,by="Interval")
stepsMerged$NewSteps <- ifelse(is.na(stepsMerged$Steps.x),stepsMerged$Steps.y,stepsMerged$Steps.x)
stepsNew <- stepsMerged[c("NewSteps","Date","Interval")]
stepsNewByDay <- aggregate(x = stepsNew$NewSteps , by = list(stepsNew$Date), FUN = sum ,na.rm=TRUE)
names(stepsNewByDay) <- c("Date","Steps")
hist(stepsNewByDay$Steps, xlab="number of steps", main="Total number of steps taken each day",col="gray")

```

## Are there differences in activity patterns between weekdays and weekends?


```{r compareweekdaysweekends }
#Get Weekend averages
weekendStepData <- stepData[weekdays(stepData$Date) %in% c("Saturday","Sunday"), ]
weekendintervalsplit <- split(weekendStepData$Steps,weekendStepData$Interval)
weekendmeanStepsInterval <- sapply(weekendintervalsplit, mean,na.rm=TRUE)

#Get Weekday averages
weekdayStepData <- stepData[!(weekdays(stepData$Date) %in% c("Saturday","Sunday")), ]
weekdayintervalsplit <- split(weekdayStepData$Steps,weekdayStepData$Interval)
weekdaymeanStepsInterval <- sapply(weekdayintervalsplit, mean,na.rm=TRUE)

plot(names(meanStepsInterval), meanStepsInterval , type = "l", 
main = "Mean steps per interval comparison Weekends and Weekdays",  xlab = "Interval", 
ylab = "Mean # of steps ",col="blue"  )
lines(names(weekdayintervalsplit), weekdaymeanStepsInterval , col="green")
lines(names(weekendintervalsplit),weekendmeanStepsInterval,col="red")
legend("topright", c("All","Weekday", "Weekend"), col = c("blue","green", "red"), lty = 1)

```

